Template for preparing your Brief Report submission to PNAS using Overleaf

### 2023-11-22 LPR/index.qmd
doesn't compile without error because of the \hh
Phrasing tweaks and updates

When an examiner conducts a toolmark assessment of wire cut evidence, they are typically working with a tool recovered from the suspect as well as a wire recovered from the scene. The examiner will typically take the tool and use it to cut a piece of metal, known as a blade cut, to record the markings of the entire cutting surface simultaneously. This process may be repeated at multiple angles in order to record the effect of different contact surfaces on the resulting striations. Markings on the wire will then be compared to these blade cuts using either visual inspection or perhaps in the future a numerical comparison algorithm similar to those developed for bullet striations [@hare2017automatic] and cartridge cases [@zemmels2023]. \hh{let's not just cite CSAFE we are not the only ones. }

@@

All of these situations may occur during white box studies (though the final situation is very improbable). As a result, perhaps the best estimate of the probability of a false identification that we can obtain is the false identification rate of one or more white box studies. We have previously published a paper detailing the error rates of different firearms and toolmark studies; to our knowledge there are no similar white box studies for wire cuts. However, if we confine ourselves to error rates for striated materials (bullets, aperture shear, ejectors, screwdrivers) we may be able to get a reasonable estimate for the error rate of wire cuts. We assembled 3 black-box error rate studies of striated toolmarks (bullets and firing pins) [@bajic2020; @mattijssen2021; @best2022]; these studies estimate the false positive rate (e.g. the number of different-source identifications relative to the total number of different-source comparisons) rate to be between 0.07235 [@mattijssen2021, firing pins only] and 0.004545 [@best2022]. While there are a number of problems with these studies (which we have detailed in previous papers, i.e. @inconclusives), our goal here is to use the best estimates we can get of false positive error rates in particular, as the false positive rate is the main driver of the problem with multiple comparisons. If we calculate a pooled false positive rate from all studies, we arrive at a false positive rate of $\frac{20 + 56 + 1}{2842 + 774 + 220} = `r sprintf('%0.02f', 77/(2842+774+220))`$, that is, about 2%.

@@

\begin{align*}
P(\text{False positive error in one comparison}) &= `r fprformat`\\
P(\text{No false positive error in one comparison}) &= `r sprintf('%0.04f', 1-fpr)`\\
P(\text{no false positive errors in } N \text{ comparisons}) &= `r sprintf('%0.04f', 1-fpr)`^N\\
P(\text{at least one false positive error in } N \text{ comparisons}) &= 1 - `r sprintf('%0.04f', 1-fpr)`^N  \tag{A}\label{eq:fpr}\\
&= `r sprintf("%.04f", 1-(1-fpr)^comparisons)` %>% 
\end{align*}

It then becomes relevant to consider how low the error rate would need to be for striated comparisons in order to make wire cut examination plausible. If we control the overall error rate amid $N$ independent comparisons, our estimated FPR for a single comparison can be calculated by solving Equation \ref{eq:fpr-control} for $r$.

@@

\begin{align*}
P(F_1) &= r\\
P(\overline{F_1}) &= 1-r\\
P(F_n) &= \left[1-r\right]^n\\
P(\overline{F_n}) &= 1 - \left[P(\overline{F_1})\right]^n = 1 - \left[1 - r\right]^n\tag{B}\label{eq:fpr-control}\\
r &= 1 - \left[1 - P(F_n)\right]^{1/n}
\end{align*}

We can explore what this means visually by considering different values of $N$ and $r$ to see how the overall false positive error rate behaves, as shown in @fig-control-overall-fpr. The astonishing conclusion is that even at the smallest error rate for striated comparisons, 0.45% [@bajic2020], we have about a 3% chance of at least one false positive conclusion when 10 comparisons are made and about a 37.5% chance of at least one false positive conclusion when 100 comparisons are made! These results suggest that under current estimates of false positive error rates, it is nearly impossible for toolmark comparisons of wires to provide compelling, "beyond a reasonable doubt" levels of evidence.

>>>>

When an examiner conducts a tool mark assessment of wire cut evidence, they are typically working with a tool recovered from the suspect as well as a wire recovered from the scene. The examiner will typically take the tool and use it to cut a piece of metal, known as a blade cut, to record the markings of the entire cutting surface simultaneously. This process may be repeated at multiple angles in order to record the effect of different contact surfaces on the resulting striations. Markings on the wire will then be compared to these blade cuts using either visual inspection or perhaps in the future a numerical comparison algorithm similar to those developed for bullet striations [@hare2017automatic; @mattijssen2021] and cartridge cases [@zemmels2023; @tong2014; @geradts2001].

@@

All of these situations may occur during black box studies (though the final situation is very improbable). As a result, perhaps the best estimate of the probability of a false identification that we can obtain is the false identification rate of one or more white box studies. We have previously published a paper detailing the error rates of different firearms and toolmark studies; to our knowledge there are no similar white box studies for wire cuts. However, if we confine ourselves to error rates for striated materials (bullets, aperture shear, ejectors, screwdrivers) we may be able to get a reasonable estimate for the error rate of wire cuts. We assembled 3 black-box error rate studies of striated toolmarks (bullets and firing pins) [@bajic2020; @mattijssen2021; @best2022]; these studies estimate the false positive rate (e.g. the number of different-source identifications relative to the total number of different-source comparisons) rate to be between 0.07235 [@mattijssen2021, firing pins only] and 0.004545 [@best2022]. While there are a number of problems with these studies (which we have detailed in previous papers, i.e. @inconclusives), our goal here is to use the best estimates we can get of false positive error rates in particular, as the false positive rate is the main driver of the problem with multiple comparisons. If we calculate a pooled false positive rate from all studies, we arrive at a false positive rate of $\frac{20 + 56 + 1}{2842 + 774 + 220} = `r sprintf('%0.02f', 77/(2842+774+220))`$, that is, about 2%.

@@

```{=tex}
\begin{align*}
P(\text{False positive error in one comparison}) &= `r fprformat`\\
P(\text{No false positive error in one comparison}) &= `r sprintf('%0.04f', 1-fpr)`\\
P(\text{no false positive errors in } N \text{ comparisons}) &= `r sprintf('%0.04f', 1-fpr)`^N\\
P(\text{at least one false positive error in } N \text{ comparisons}) &= 1 - `r sprintf('%0.04f', 1-fpr)`^N  \tag{A}\label{eq:fpr}\\
&= `r sprintf("%.04f", 1-(1-fpr)^comparisons)` %>% 
\end{align*}
```
It then becomes relevant to consider how low the error rate would need to be for striated comparisons in order to make wire cut examination plausible. If we control the overall error rate amid $N$ independent comparisons, our estimated FPR for a single comparison can be calculated by solving Equation \ref{eq:fpr-control} for $r$.

@@

```{=tex}
\begin{align*}
P(F_1) &= r\\
P(\overline{F_1}) &= 1-r\\
P(F_n) &= \left[1-r\right]^n\\
P(\overline{F_n}) &= 1 - \left[P(\overline{F_1})\right]^n = 1 - \left[1 - r\right]^n\tag{B}\label{eq:fpr-control}\\
r &= 1 - \left[1 - P(F_n)\right]^{1/n}
\end{align*}
```
We can explore what this means visually by considering different values of $N$ and $r$ to see how the overall false positive error rate behaves, as shown in @fig-control-overall-fpr. The astonishing conclusion is that even at the smallest error rate for striated comparisons, 0.45% [@best2022], we have about a 3% chance of at least one false positive conclusion when 10 comparisons are made and about a 37.5% chance of at least one false positive conclusion when 100 comparisons are made! These results suggest that under current estimates of false positive error rates, it is nearly impossible for toolmark comparisons of wires to provide compelling, "beyond a reasonable doubt" levels of evidence.

### 2023-11-22 PNAS/index.Rnw
Phrasing tweaks and updates

\documentclass[9pt,twocolumn,twoside]{pnas-new}
% Use the lineno option to display guide line numbers if required.

\templatetype{pnasbriefreport} % Choose template
% {pnasresearcharticle} = Template for a two-column research article
% {pnasmathematics} %= Template for a one-column mathematics article
% {pnasinvited} %= Template for a PNAS invited submission

@@

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}

@@

\dropcap{M}ultiple comparisons are a persistent problem for statistical methods and for science more generally. XXX cite XXX In forensic statistics, we have been focusing on the necessity of building up research databases which can be used to develop objective algorithms for evidence comparisons and explore closest non-matches as suggested in \cite{pcast}. While these databases (and open data more generally) are critically important, we also need to be careful about how our methods are affected by the multiple comparison problem.

As we develop objective algorithms and standardize laboratory procedures in forensic pattern evidence, we need to be aware of the fact that we often are conducting many comparisons, either implicitly or explicitly. Algorithms in pattern evidence often utilize cross-correlation \cite{vorburgerApplicationsCrosscorrelationFunctions2011} to align different data sources, implicitly making comparisons along the entire length or surface of the data. Similarly, forensic examiners visually rotate or align two pieces of evidence under comparison microscopes. When we conduct database searches, these searches often return the $N$ best matches from the database, but this requires making comparisons to all of the database entries which meet the search criteria.

@@

@@

\matmethods{Please describe your materials and methods here. This can be more than one paragraph, and may contain subsections and equations as required.

\subsection*{Subsection for Method}
Example text for subsection.
}

\showmatmethods{} % Display the Materials and Methods section

@@

\bibsplit[3]
% %Use \bibsplit to split the references from the body of the text. Value "[3]" represents the number of reference in the left column (Note: Please avoid single column figures & tables on this page.)

>>>>

\documentclass[9pt,twocolumn,twoside]{pnas-new}
% Use the lineno option to display guide line numbers if required.
\usepackage{cleveref}
\templatetype{pnasbriefreport} % Choose template
% {pnasresearcharticle} = Template for a two-column research article
% {pnasmathematics} %= Template for a one-column mathematics article
% {pnasinvited} %= Template for a PNAS invited submission

@@

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
<<setup, include = F>>=
library(knitr)
opts_chunk$set(error=F, warning=F, message = F, dpi=300, echo = F)
@

@@

\dropcap{M}ultiple comparisons are a persistent problem for statistical methods and for science. In forensic statistics, researchers have been focusing on the necessity of building up research databases which can be used to develop objective algorithms for evidence comparisons and explore closest non-matches \citep{pcast}. While these databases (and open data) are critically important in forensics, we also need to be careful about how our methods are affected by the multiple comparison problem.

As we develop objective algorithms and standardize laboratory procedures in forensic pattern evidence, we must remain cognizant that we often are conducting many comparisons, either implicitly or explicitly. Algorithms in pattern evidence often utilize cross-correlation \citep{vorburgerApplicationsCrosscorrelationFunctions2011} to align different data sources, implicitly making comparisons along the entire length or surface of the data. Similarly, forensic examiners visually rotate or align two pieces of evidence under comparison microscopes. When we conduct database searches, these searches often return the $N$ best matches from the database, but this requires making comparisons to all of the database entries which meet the search criteria.

@@

\section*{Probability of False Identifications}

There are two components to a false positive rate: coincidental matches (that is, two pieces of evidence that happen to have similar characteristics but are not from the same source) and human/technical failures \citep[p 50]{pcast}. In objective disciplines with matching rules, such as DNA, these sources can be distinguished, and when coincidental match rates are very low, human/technical failures make up a large component of the error rate. However, in toolmark examination, we do not have an objective matching rule; examiners testify based on subjective, individual rules for how much similarity is enough for an identification. As a result, we cannot separate coincidental match rates from human/technical failures.
<<fpr, include = F>>=
errors <- c(20, 56, 1)
knm <- c(2842, 774, 220)

fpr <- sum(errors)/sum(knm)
@

Compounding this problem, open-set studies of striated comparisons (bullets, firing pin impressions) \citep{bajic2020, mattijssen2021,best2022} suggest the false positive rate for these comparisons is between 0.004545\citep{best2022} and 0.07235\citep{mattijssen2021}; pooling data from these three studies yields a false positive rate of is \Sexpr{sprintf("%.02f%%", fpr*100)}, which is high relative to the consequences of a false identification of evidence from different sources.

If we begin by considering the multiple comparison problem relative to the $\ell/d$ independent comparisons between a blade cut and a wire, we can calculate the overall false positive rate for all comparisons using $r$, the false positive rate for one comparison, and simple probability. Let $F_n:=\{\text{at least one false positive in }n\text{ comparisons}\}$. Then,
\begin{align}
P(F_1) &= r\nonumber\\
P(\overline{F_1}) &= 1-r\nonumber\\
P(F_n) &= \left[1-r\right]^n\nonumber\\
P(\overline{F_n}) &= 1 - \left[P(\overline{F_1})\right]^n = 1 - \left[1 - r\right]^n.\label{eq:fpr-control}
\end{align}
Using this relationship between $F_n$ and the false positive rate $r$, we can determine what $r$ must be in order to limit $P(F_n)$ to an acceptable level by solving for $r$: $r = 1 - \left[1 - P(F_n)\right]^{1/n}$.

\Cref{fig:control-overall-fpr} explores the implications of this relationship. The astonishing conclusion is that even at the smallest error rate for striated comparisons, we have about a 3\% chance of at least one false positive conclusion when 10 comparisons are made and about a 37.5\% chance of at least one false positive conclusion when 100 comparisons are made. These results suggest that using current estimates of false positive error rates, it is nearly impossible for wire comparisons to provide compelling, 'beyond a reasonable doubt' levels of evidence.
<<control-overall-fpr, fig.cap="Exploration of overall error rate estimates for different values of $r$, the false positive rate for a single comparison. Dotted vertical lines show false positive error rate estimates (from left to right, \\citep{best2022,bajic2020,mattijssen2021}).", fig.width = 8, fig.height = 4>>=
library(ggplot2)
library(tidyr)
library(dplyr)
library(latex2exp)
library(geomtextpath)
single_rmp = exp(seq(-1000, 0, .1))
N = c(10, 100, 1000, 10000)

res = crossing(single_rmp, N) %>%
  mutate(overall = 1 - (1 - single_rmp)^N)

studies <- tibble(single_rmp = c( 0.004545,0.00704, 0.0724), study = c("Best (2022)","Bajic (2020)",  "Mattijssen (2021)"), y = c(-0.025, .15, 0.6), hjust = c(0, 0, 1))
ggplot(res, aes(x = single_rmp, y = overall, group = factor(N), color = factor(N))) +
  geom_line() +
  coord_cartesian(xlim = c(0, 0.07), ylim = c(0, 1)) +
  ylab("Probability of a coincidental match (N comparisons)") +
  xlab("Probability of a false positive (1 comparison)") +
  scale_y_continuous(breaks = seq(0, 1, .1), minor_breaks = seq(0, 1, .05)) +
  ggtitle("Wire Comparison: Controlling Overall False Positive Rate") +
  geom_vline(xintercept = 0.02, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = studies$single_rmp, linetype = "51", color = "grey50") +
  geom_label(data = studies[1,], aes(x = single_rmp, y = y, label = study), inherit.aes = F, hjust = studies$hjust[1], vjust = 0) +
  geom_label(data = studies[2,], aes(x = single_rmp, y = y, label = study), inherit.aes = F, hjust = studies$hjust[2], vjust = 0) +
  geom_label(data = studies[3,], aes(x = single_rmp, y = y, label = study), inherit.aes = F, hjust = studies$hjust[3], vjust = 0) +
  annotate(geom = "label", x = 0.02, y = 0.1, label = "Pooled error rate \nused in this paper", color = "grey40", hjust=0) +
  scale_color_discrete("# comparisons") +
  theme_bw() +
  theme(legend.position = c(1, 0), legend.justification = c(1, 0), legend.background = element_rect(color = "black"))
@

If we look at this data differently, we can assess the probability that $N$ comparisons are made accurately (e.g. there are no coincidental matches). It is clear from \Cref{fig:Accuracy} that with existing false positive error rate estimates between 0.07 and 0.0045, the accuracy of the aggregate comparison between a wire and a tool becomes questionable, as the number of candidate alignments that must be examined is too high. Even the most innocuous example we could come up with (small blade, only 2 cutting surfaces, and a relatively large wire) involved a minimum of \Sexpr{2*ncomp_min} completely independent comparisons. As a result, it seems improbable that a wire comparison made under current protocols (with corresponding error rates) would have an overall accuracy rate which would be reliable enough to be presented at trial. When we then consider that examiners would make cuts at multiple angles \citep{baikerToolmarkVariabilityQuality2015}, increasing the number of comparisons even in the simple case we have presented here, the probability of a coincidental match increases

<<Accuracy, fig.cap="Probability of accurate decisions for $N$ comparisons for different error rates.", fig.width = 8, fig.height = 4>>=
r_vals <- tibble(r = c(.0005, 0.001, 0.005, .01, 0.05),
                 rlab = sprintf("P(F[1])==%0.05f", r))
res2 <- crossing(r = r_vals$r,
                 N = c(seq(1, 11, .01),seq(11.1, 109.9, .1), seq(110, 1100, 1), seq(1110, 10000, 10))) %>%
  mutate(overall = (1 - r)^N) %>%
  left_join(r_vals)

breakpt <- .9

error_res <- filter(res2, lead(overall, 1) <= breakpt)
res2 <- filter(res2, overall >= breakpt)

res2sum <- res2 %>% group_by(r) %>% filter(overall == min(overall))

ggplot(data = res2, aes(x = N, y = overall, color = factor(r), label = rlab)) +
  geom_textline(parse = T, gap = F, straight = F, padding = unit(1, "mm"), vjust = 0, linewidth = 0.75) +
  geom_line(data = error_res, linetype = 2, linewidth = 0.75) +
  xlab("Number of comparisons") +
  scale_x_log10() +
  geom_hline(yintercept = breakpt, linewidth = 1) +
  geom_label(data = res2sum, aes(x = N, y = breakpt, label = paste0("N == ", round(N))), parse = T, vjust = 1, hjust = 1) +
  ylab("Probability of no coincidental matches (N comparisons)") +
  ggtitle("# Comparisons and Overall Accuracy") +
  scale_color_discrete("FP Rate for single comparison", guide = 'none') +
  theme_bw() +
  coord_cartesian(ylim = c(0.75, 1.0)) +
  theme(axis.text.y = element_text(hjust = 0.5, vjust = 0.5, angle = 90))

@

@@

% \matmethods{Please describe your materials and methods here. This can be more than one paragraph, and may contain subsections and equations as required.
% \subsection*{Subsection for Method}
% Example text for subsection.
% }
% \showmatmethods{} % Display the Materials and Methods section

@@

% \bibsplit[3]
% %Use \bibsplit to split the references from the body of the text. Value "[3]" represents the number of reference in the left column (Note: Please avoid single column figures & tables on this page.)
